A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-11-04 hybrid-nts.tex
Add 3-D results content

I ran all cases on OpenMC to generate the required group constants and reference $k_\text{eff}$
values to assess the accuracy of the deterministic methods ($S_N$, neutron diffusion, hybrid 
methods) against OpenMC in continuous-energy (OpenMC-CE) and multigroup (OpenMC-MG) modes. Table
\ref{table:var} details how OpenMC and the deterministic methods represent the position, direction
of travel, neutron energy, and angle-dependence in $\Sigma_s$. The hybrid $S_N$-diffusion method
features representations from either $S_N$ or neutron diffusion depending on the subregion under
consideration. Comparing OpenMC-MG results with OpenMC-CE results allows us
to quantify errors arising from neutron energy group discretization and the scattering cross
section simplifications.

@@

\subsection{2-D Neutronics Model Setup}

@@

\begin{itemize}
  \item Ignored thermal expansion of all reactor components.
  \item Modeled the control rods as perfectly annular cylinders instead of beaded control elements
    strung on flexible steel cables. Ignored the inconel shells encasing the control elements and
    the steel helical cables linking the control elements together.
  \item Set channel pitch, width, and length to 2, 0.2, and 1.2 inches, respectively.
  \item Homogenized the INOR, graphite, and molten salt regions in the sample basket.
  \item Modified the outer edge of the core graphite to be circular with no jagged edges.
  \item Replaced partial channels with full-size channels near the edge of the core graphite.
  \item Ignored the thermal insulation and shield structures outside the core.
\end{itemize}

@@

\subsubsection{2-D quarter-core results}

@@

\subsubsection{2-D full-core results}

@@

TBD

%\subsection{3-D Model Setup}

%\subsection{3-D Numerical Results}

@@

TBD

>>>>

I ran all cases on OpenMC to generate the required group constants and reference $k_\text{eff}$
values to assess the accuracy of the deterministic methods ($S_N$, neutron diffusion, hybrid 
methods) against OpenMC in continuous-energy (OpenMC-CE) and multigroup (OpenMC-MG) modes. Table
\ref{table:var} details how OpenMC and the deterministic methods represent the position, direction
of travel, neutron energy, and angle-dependence in $\Sigma_s$. The hybrid $S_N$-diffusion method
features representations from either $S_N$ or neutron diffusion depending on the subregion under
consideration. Comparing OpenMC-MG results with OpenMC-CE results allows us
to quantify errors arising from neutron energy group discretization and the scattering cross
section approximations.

@@

\subsection{2-D Neutronics Model Setup} \label{sec:2d-model-setup}

@@

\begin{itemize}
  \item Ignored thermal expansion of all reactor components.
  \item Modeled the control rods as perfectly annular cylinders instead of beaded control elements
    strung on flexible steel cables. Ignored the inconel shells encasing the control elements and
    the steel helical cables linking the control elements together.
  \item Set channel pitch, width, and length to 2, 0.2, and 1.2 inches, respectively.
  \item Homogenized the INOR-8, graphite, and molten salt regions in the sample basket.
  \item Modified the outer edge of the core graphite to be circular with no jagged edges.
  \item Replaced partial channels with full-size channels near the edge of the core graphite.
  \item Ignored the thermal insulation and shield structures outside the core.
\end{itemize}

@@

\subsubsection{2-D Quarter-Core}

@@

\subsubsection{2-D Full-Core}

@@

\subsection{3-D Neutronics Model Setup} \label{sec:3d-model-setup}

Figure \ref{fig:msre-picture} shows the vertical cross section of the actual \gls{MSRE} vessel.
The actual design has rounded upper and lower plena to help direct salt flow through the system.
This work uses a modified 3-D numerical \gls{MSRE} model as shown by the vertical cross section
in Figure \ref{fig:msre-geom-vert}. The model excludes the toroidal flow distributor near the top
of the vessel and reshapes the plena into right cylinders. It has the same horizontal cross section
as the 2-D model (Figure \ref{fig:full-geom}) throughout most of the salt-graphite lattice. In
addition to modifications in the 2-D model listed in Section \ref{sec:2d-model-setup}, the
3-D model includes the following modifications:

\begin{itemize}
  \item Removed the toroidal flow distributor near the top of the vessel.
  \item Reshaped the upper and lower plena into right cylinders while keeping the salt volume
    constant.
  \item Flattened the conical tops of the graphite lattice blocks.
  \item Regularized the complicated graphite lattice structure at the bottom.
  \item Set a fixed outer vessel thickness of 2.54 cm.
  \item Homogenized the lower plenum by 90.8\% salt- 9.2\% INOR-8 (vessel) by volume.
  \item Removed the reactor access nozzle and the various components within it.
  \item Added a graphite block at the bottom of the control rod thimbles.
\end{itemize}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.97\columnwidth]{msre-picture}
  \caption{Vertical cross section of the actual \gls{MSRE} vessel. It has rounded upper and lower
  plena, and a toroidal flow distributor that distributes the salt inflow evenly as it enters the
  core through holes on the inside vessel surface of the distributor.}
  \label{fig:msre-picture}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=\columnwidth]{msre-geom-vert}
  \caption{Vertical cross section of the 3-D numerical \gls{MSRE} model.}
  \label{fig:msre-geom-vert}
\end{figure}

Table \ref{table:reactor-dimensions} lists relevant dimensions of the 3-D \gls{MSRE} model.
All material compositions remain
the same as the 2-D models and the \gls{MSRE} benchmark specifications \cite{fratoni_molten_2020}.
All 3-D neutronics simulations in this section used the molten salt composition with 65.25 kg
$^{235}$U loaded (Table \ref{table:2d-salt-composition}), except the simulations for temperature
reactivity coefficient calculations which had a higher $^{235}$U loading of 67.00 kg (Table
\ref{table:67-salt-composition}.

\begin{table}[htb]
  \centering
  \caption{List of reactor dimensions.}
  \label{table:reactor-dimensions}
\end{table}

\begin{table}[htb]
  \small
  \centering
  \setlength\tabcolsep{4pt}
  \caption{\gls{MSRE} molten salt composition when the $^{235}$U loading was at 65.25 kg.}
  \begin{tabular}{l S S S S S S S}
    \toprule
    Element & {Li} & {Be} & {Zr} & {Hf} & {$^{234}$U} & {$^{235}$U} & {$^{236}$U} \\
    \midrule
    Composition [kg] & 507.42 & 293.96 & 513.97 & 0.0029 & 0.68 & 67.00 & 0.28 \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{l S S S S S S}
    \toprule
    Element & {$^{238}$U} & {Fe} & {Cr} & {Ni} & {O} & {F} \\
    \midrule
    Composition [kg] & 142.02 & 0.75 & 0.13 & 0.14 & 2.27 & 3104.22 \\
    \bottomrule
  \end{tabular}
  \label{table:67-salt-composition}
\end{table}

All neutronics simulations ran on OpenMC under continuous energy mode and on Moltres with the
hybrid $S_N$-diffusion method. The hybrid simulations ran with a maximum stabilization factor of
$c=250$, a void constant $\varsigma=0.5$, and a uncorrected diffusion coefficient cap of $D_g=2.5$.
Refer to Sections \ref{sec:void-treatment} and \ref{sec:diffcoef-cap} for the relevant discussion
on void treatment for near-void regions and the impact of large diffusion coefficient values on
solver performance. As for the standard neutron diffusion method, it either converges very slowly
or does not converge at all when $D_g$ values are uncapped due to excessive diffusive effects in
the air-filled control rod thimbles. Group-wise $D_g$ values, generated using OpenMC, ranged from
100 cm to 850 cm in the air regions as compared to the typical 0.5 cm to 2.0 cm range observed in
non-air regions. Therefore, all neutron diffusion method simulation results in the following
section were measured with the diffusion coefficient capped at $D_g=2.5$.

Unlike the 2-D hybrid simulations which ran with $S_8$ angular discretization, the 3-D hybrid
simulations ran with $S_6$ angular discretization due to significantly
larger computational and memory requirements demanded by the 3-D simulations. Further discussion on
computational performance may be found at the end of Section \ref{sec:3d-nts}. All simulations,
except those for temperature reactivity coefficient calculations, ran on isothermal 3-D models at
911 K.

\subsection{3-D Numerical Results} \label{sec:3d-nts}

\subsubsection{Initial Criticality Configuration}

The \gls{MSRE} achieved initial criticality with Rod 1 inserted to a height of 46.6 inches relative
to the fully inserted position. For reference, the rods are at a height of 51 inches when fully
withdrawn. Table \ref{table:initial-crit}
shows the $k_\text{eff}$ values from \gls{MSRE} experimental data, numerical benchmark data
\cite{fratoni_molten_2020}, and this work. All numerical models overestimate the $k_\text{eff}$
relative to the experimental value. In the \gls{MSRE} numerical benchmark report, the authors
attribute this $k_\text{eff}$ discrepancy largely to discrepancies in the carbon cross section data
in nuclear data libraries. They noted similar $k_\text{eff}$ overestimations by 1-2\% in benchmark
results for the HTTR, a graphite-moderated, helium-cooled reactor. The OpenMC and Moltres
simulations in this work used the same
ENDF/B-VII.1 nuclear data library as the \gls{MSRE} numerical benchmark report.

\begin{table}[htb]
  \centering
  \caption{$k_\text{eff}$ values from \gls{MSRE} experimental data, the \gls{MSRE} numerical
  benchmark \cite{fratoni_molten_2020}, and the OpenMC and Moltres models in this work.}
  \begin{tabular}{l S[table-format=1.5(2)]}
    \toprule
     & {$k_\text{eff}$} \\
     \cmidrule(l){1-1}
    \gls{MSRE} experimental data & 1.00000(420) \\
    Serpent 2 (Numerical benchmark) & 1.02132(3) \\
    OpenMC (This work) & 1.01308(20) \\
    Hybrid (This work) & 1.01955 \\
    Diffusion (This work) & 1.01885 \\
    \bottomrule
  \end{tabular}
  \label{table:initial-crit}
\end{table}

Of the four $k_\text{eff}$ values, we expect to see the closest agreement between the Serpent 2
model from the numerical benchmark and OpenMC from this work. However, the OpenMC \gls{MSRE} model
differs from the Serpent 2 \gls{MSRE} models due to geometry modifications listed in Sections
\ref{sec:2d-model-setup} and \ref{sec:3d-model-setup}. The benchmark report quantified the impact
of several geometry modifications, namely the removal of the flow distributor, the homogenization
of the sample basket, and the removal of the thermal shield. These modifications contribute to
-$0.098\%$, -$0.037\%$, and -$0.885\%$ change in the $k_\text{eff}$, respectively, for a final
$k_\text{eff}$ estimate of $1.01093(3)$. This brings the OpenMC $k_\text{eff}$ estimate of
$1.01308(20)$ to closer agreement with the numerical benchmark value.

The hybrid method $k_\text{eff}$ estimate is 0.00647 higher than the OpenMC estimate. For this
initial criticality configuration, the neutron diffusion method $k_\text{eff}$ estimate is closer
to OpenMC than the hybrid method. Overall, the OpenMC and Moltres models show good agreement with
the Serpent 2 model from the numerical benchmark.
Figures * show the neutron group fluxes on the interior of the 3-D models. The salt-graphite
lattice structure is visible from the group flux distributions. Group 7 and 8 neutron fluxes are
particularly small near the control rods due to strong thermal neutron capture by Gadolinium in the
rods.

\subsubsection{Control Rod Worth}

After achieving initial criticality, \gls{ORNL} researchers ran control rod calibration experiments
with the \gls{MSRE} at zero-power (typically at about 10 W power output).

For control rod worth calculations, the OpenMC and Moltres simulations ran  with Rod 1 at fourteen
different rod positions from full insertion (0 inch) to full withdrawal (51 inches). The rod
positional limits of 0 inch and 51 inches represents the full range of deliberate rod motion
allowable by the control rod drive motor \cite{robertson_msre_1965}.

\begin{figure}[p]
  \centering
  \includegraphics[width=0.8\columnwidth]{rod-worth-curve}
  \caption{Reactivity inserted by Rod 1 at various rod positions relative to the full insertion.}
  \label{fig:rod-worth}
\end{figure}

\subsubsection{Temperature Reactivity Coefficient}

\subsubsection{Computational Performance}

@@

### 2024-11-04 hybrid.tex
Add 3-D results content

\subsection{Spatial Discretization and Void Treatment}

@@

We can now proceed with the void treatment for stabilization.
We start by defining the following stabilization operator
%
\begin{gather}
  \bm{\tau} \equiv
  \begin{bmatrix}
    \tau_1 \\
    \tau_2 \\
    \vdots \\
    \tau_G
  \end{bmatrix}
  \shortintertext{where}
  \begin{align*}
    \tau_g &=
    \begin{cases}
      \frac{1}{\Sigma_{t,g}} \mbox{ for } h\Sigma_{t,g} \geq \varsigma \\
      \frac{h}{\varsigma} \mbox{ for } h\Sigma_{t,g} < \varsigma
    \end{cases}, & \\
    h &= \mbox{mesh element size,} & \\
    \varsigma &= \mbox{void constant.} &
  \end{align*}
\end{gather}
%
Next, we redefine Eq.\ \ref{eq:afe} as
%
\begin{gather}
  \bm{\Psi} = \left(\mathbb{I}-\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}+
  \bm{\tau}\left[\mathbb{S}\bm{\Psi}+\bm{Q}
  -\frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)-\mathbb{L}_1\bm{\Psi}\right]
  \label{eq:afe-vt}
\end{gather}
%
Substituting Eq.\ \ref{eq:afe-vt} into Eq.\ \ref{eq:nte-weak} yields the stabilized \gls{SAAF}
formulation given as
%
\begin{multline}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,
  \frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)\right) + 
  \left(\mathbb{L}_1\bm{\Psi}^*,
  \left(\bm{\tau}\mathbb{L}_1-\mathbb{I}+\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}\right) +
  \langle\bm{\Psi}^*,\bm{\Psi}\rangle^+ - \langle\bm{\Psi}^*,\bm{\Psi}\rangle^- +
  \left(\mathbb{L}_2\bm{\Psi}^*,\bm{\Psi}\right) \\
  = \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{S}\bm{\Psi}\right) +
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\bm{Q}\right)
  \label{eq:saaf-vt}
\end{multline}
%
In non-void regions where $\Sigma_{t,g}$ is non-zero, we can disable the stabilization scheme by
setting $\varsigma$ to zero. Then $\bm{\tau}\mathbb{L}=\mathbb{I}$
and the stabilized \gls{SAAF} formulation reduces to the unstabilized formulation in Eq.\
\ref{eq:saaf}.

@@

>>>>

\subsection{Spatial Discretization and Void Treatment} \ref{sec:void-treatment}

@@

We can now proceed with the void treatment for stabilization.
We start by defining the following stabilization operator
%
\begin{gather}
  \bm{\tau} \equiv
  \begin{bmatrix}
    \tau_1 \\
    \tau_2 \\
    \vdots \\
    \tau_G
  \end{bmatrix}
  \shortintertext{where}
  \begin{align*}
    \tau_g &=
    \begin{cases}
      \frac{1}{c\Sigma_{t,g}} \mbox{ for } ch\Sigma_{t,g} \geq \varsigma \\
      \frac{h}{\varsigma} \mbox{ for } ch\Sigma_{t,g} < \varsigma
    \end{cases}, & \\
    h &= \mbox{mesh element size,} & \\
    c &= \mbox{maximum stabilization factor,} & \\
    \varsigma &= \mbox{void constant.} &
  \end{align*}
\end{gather}
%
Next, we redefine Eq.\ \ref{eq:afe} as
%
\begin{gather}
  \bm{\Psi} = \left(\mathbb{I}-\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}+
  \bm{\tau}\left[\mathbb{S}\bm{\Psi}+\bm{Q}
  -\frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)-\mathbb{L}_1\bm{\Psi}\right]
  \label{eq:afe-vt}
\end{gather}
%
Substituting Eq.\ \ref{eq:afe-vt} into Eq.\ \ref{eq:nte-weak} yields the stabilized \gls{SAAF}
formulation given as
%
\begin{multline}
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,
  \frac{\partial}{\partial t}\left(\frac{\bm{\Psi}}{\bm{v}}\right)\right) + 
  \left(\mathbb{L}_1\bm{\Psi}^*,
  \left(\bm{\tau}\mathbb{L}_1-\mathbb{I}+\bm{\tau}\mathbb{L}_2\right)\bm{\Psi}\right) +
  \langle\bm{\Psi}^*,\bm{\Psi}\rangle^+ - \langle\bm{\Psi}^*,\bm{\Psi}\rangle^- +
  \left(\mathbb{L}_2\bm{\Psi}^*,\bm{\Psi}\right) \\
  = \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\mathbb{S}\bm{\Psi}\right) +
  \left(\left(\mathbb{I}+\bm{\tau}\mathbb{L}_1\right)\bm{\Psi}^*,\bm{Q}\right)
  \label{eq:saaf-vt}
\end{multline}
%
In non-void regions where $\Sigma_{t,g}$ is non-zero, we can disable the stabilization scheme by
setting $\varsigma$ to zero. Then $\bm{\tau}\mathbb{L}=\mathbb{I}$
and the stabilized \gls{SAAF} formulation reduces to the unstabilized formulation in Eq.\
\ref{eq:saaf}.

@@

\subsection{Setting Caps on the Diffusion Coefficient Values} \ref{sec:diffcoef-cap}

From experience, the $S_N$-diffusion outer iterations under both hybrid schemes or the standard
$S_N$ method with nonlinear diffusion acceleration converged at a faster rate when the uncorrected
diffusion coefficient values are capped at around $D_g=2.5$ cm. Excessively large $D_g$ values,
which occur in near-void regions, cause significant computational performance degradation. The caps
have no impact on any method because the diffusion term containing $D_g$ is canceled out by the
transport correction terms as the solution converges. This behavior was also confirmed through
email correspondence with Dr.\ Yaqi Wang regarding nonlinear diffusion acceleration of the
\gls{SAAF} $S_N$ method \cite{wang_diffusion_2014}.
